---
title: "Getting started with SDALGCP2 (simulated example)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with SDALGCP2 (simulated example)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`SDALGCP2` fits a spatially discrete approximation to a log-Gaussian Cox process
(SDA-LGCP) to spatially aggregated disease counts. It is a faster, modernised
successor to `SDALGCP`: the heavy steps (aggregated correlation assembly, the
MALA sampler, the Monte Carlo likelihood) run in C++.

This vignette uses **entirely simulated data**, so it is fully reproducible
without any external files.

## 1. Simulate a lattice of regions and aggregated counts

```{r}
library(SDALGCP2)
library(sf)

set.seed(1)
bound <- st_as_sfc(st_bbox(c(xmin = 0, ymin = 0, xmax = 20, ymax = 20)))
shp   <- st_sf(geometry = st_make_grid(bound, n = c(8, 8)))   # 64 square regions
N     <- nrow(shp)

# candidate points inside each region (regular grid)
pts      <- sda_points(shp, delta = 1.2, method = 3)
phi_grid <- seq(1, 5, length.out = 8)
corr     <- precompute_corr(pts, phi_grid)

# true parameters and a latent field drawn at phi = 2.5
Sig   <- 0.5 * corr$R[, , which.min(abs(phi_grid - 2.5))]
x1    <- as.numeric(scale(st_coordinates(st_centroid(shp))[, 1]))
pop   <- round(runif(N, 500, 3000))
y     <- rpois(N, pop * exp(cbind(1, x1) %*% c(-6, 0.5) +
                            as.numeric(t(chol(Sig)) %*% rnorm(N))))
dat   <- data.frame(y = y, x1 = x1, pop = pop)
```

## 2. Fit the model

A single call generates points, builds the correlation array and runs MCML:

```{r}
ctrl <- control_mcmc(n.sim = 6000, burnin = 1500, thin = 6, h = 1.65 / N^(1/6))
fit  <- SDALGCP2(y ~ x1 + offset(log(pop)), dat, shp,
                 delta = 1.2, phi = phi_grid, method = 3, control.mcmc = ctrl)

summary(fit)
confint(fit)
```

## 3. Predict relative risk

Discrete (region-level) and continuous (change-of-support) prediction. The
`sampler = "laplace"` option skips MCMC for a near-instant approximation.

```{r}
pred_d <- predict(fit, type = "discrete",   sampler = "mcmc",    control.mcmc = ctrl)
pred_c <- predict(fit, type = "continuous", sampler = "laplace", cellsize = 1,
                  control.mcmc = ctrl)

# exceedance probability of covariate-adjusted relative risk above 1.5
exc <- exceedance(pred_d, thresholds = 1.5)
```

## 4. Speed vs the original SDALGCP

On this 64-region example the full pipeline (points + correlation + MCML) is
several times faster than `SDALGCP::SDALGCPMCML` while returning the same
estimates. The reproducible benchmark lives in `scripts/compare_vs_SDALGCP.R`:

```{r}
# source(system.file("..", "scripts", "compare_vs_SDALGCP.R", package = "SDALGCP2"))
```

Representative output (your timings will vary):

```
True:      beta=(-6, 0.5)      sigma2=0.5    phi=2.5
SDALGCP2:  beta=(-5.738,0.474) sigma2=0.687  phi=1     [1.0s]
SDALGCP:   beta=(-5.738,0.474) sigma2=0.687  phi=1     [8.4s]
Full-pipeline speedup: ~8x
```
```
